Equalizer\\Programming and User Guide

### 2011-04-04 Developer/ProgrammingGuide/paper.tex
Programming Guide: add focus distance section
[ ] May break build
[ ] May break existing applications (see CHANGES.txt)
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.8.1 for Equalizer 1.0 \\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    1.8.1   & Mar 25, 2011 & Minor Revision\\
    \hline
  \end{tabularx}}
\clearpage

@@

{\footnotesize\lstinputlisting[firstline=107, lastline=123]
  {../../../src/examples/eqHello/hello.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=57, lastline=66]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=40, lastline=55]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=67, lastline=70]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=71, lastline=78]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=80, lastline=81]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=83]
  {../../../src/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=80, lastline=88]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=90, lastline=100]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=101, lastline=118]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=120, lastline=147]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=148, lastline=152]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=154, lastline=165]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=167, lastline=175]
  {../../../src/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=61, lastline=74]
  {../../../src/examples/eqPly/initData.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=122, lastline=129]
  {../../../src/examples/eqPly/frameData.h}}
{\footnotesize\lstinputlisting[firstline=54, lastline=80]
  {../../../src/examples/eqPly/frameData.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=143, lastline=174]
  {../../../src/examples/eqPly/vertexBufferDist.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=74, lastline=95]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=97, lastline=99]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=100, lastline=118]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=130, lastline=139]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=310, lastline=317]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=319, lastline=327]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=329, lastline=359]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=387, lastline=390]
  {../../../src/examples/eqPly/config.cpp}}
{\footnotesize\lstinputlisting[firstline=450, lastline=481]
  {../../../src/examples/eqPly/config.cpp}}
{\footnotesize\lstinputlisting[firstline=528, lastline=534]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=284, lastline=308]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=790, lastline=859]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=35, lastline=48]
  {../../../src/examples/eqPly/node.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=257, lastline=268]
  {../../../src/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=57, lastline=70]
  {../../../src/examples/eqPly/pipe.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=72, lastline=78]
  {../../../src/examples/eqPly/pipe.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=39, lastline=55]
  {../../../src/examples/eqPly/pipe.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=80, lastline=84]
  {../../../src/examples/eqPly/pipe.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=76, lastline=98]
  {../../../src/examples/eqPly/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=100, lastline=109]
  {../../../src/examples/eqPly/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=69, lastline=78]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=123, lastline=137]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=139, lastline=165]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=167, lastline=178]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=591, lastline=610]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=619, lastline=689]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=238, lastline=261]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=36, lastline=57]
  {../../../src/examples/eqPixelBench/configEvent.h}}

@@

{\footnotesize\lstinputlisting[firstline=58, lastline=71]
  {../../../src/examples/eqPixelBench/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=55, lastline=61]
  {../../../src/examples/eVolve/main.cpp}}
...
{\footnotesize\lstinputlisting[firstline=92, lastline=95]
  {../../../src/examples/eVolve/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=36, lastline=44]
  {../../../src/examples/eVolve/error.h}}

@@

{\footnotesize\lstinputlisting[firstline=45, lastline=57]
  {../../../src/examples/eVolve/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=37, lastline=73]
  {../../../src/examples/eVolve/error.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=37, lastline=43]
  {../../../src/examples/eVolve/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=124, lastline=140]
  {../../../src/libs/client/aglWindow.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=67, lastline=101]
  {../../../src/libs/client/glXWindow.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=178, lastline=224]
  {../../../src/libs/client/wglWindow.cpp}}

\subsection{\label{sTracking}Head Tracking}

The \textsf{eqPly} example contains rudimentary support for head
tracking to show how head tracking can be integrated with
Equalizer. Support for a wide range of tracking devices is not within
the scope of Equalizer. Other open source and commercial implementations
cover this functionality sufficiently and can easily be integrated with
Equalizer.

An Equalizer configuration has a vector of observers. Each observer has
its own head matrix and eye separation for tracking. In \textsf{eqPly},
all observers are updated with the same head matrix. A typical
application would assign different tracking data to different observers.

\begin{figure}[h!t]\center
  \subfigure[]{\label{fMono}
    \includegraphics[height=4cm]{images/mono}}\hfil
  \subfigure[]{\label{fStereo}
    \includegraphics[height=4cm]{images/stereo}}
  \subfigure[]{\label{fTracked}
    \includegraphics[height=4cm]{images/tracked}}\hfil
  \subfigure[]{\label{fHMD}
    \includegraphics[height=4cm]{images/hmd}}
  {\caption{\label{fImmersive}Monoscopic (a), Stereoscopic (b), Tracked (c) and
      HMD (d) Frusta}}
\end{figure}

@@

In stereo rendering, the scene is rendered twice, with the two frusta
'moved' by the distance between the eyes, as shown in \fig{fStereo}.

In immersive visualization, the observer is tracked in and the view
frusta are adapted to the viewer's position and orientation, as shown
in \fig{fTracked}. The transformation $origin \rightarrow viewer$ is set by
the application using \textsf{Config::setHead\-Matrix}, which is used by
the server to compute the frusta. The resulting off-axis frusta are
positioned using the channel's head transformation, which can be
retrieved using \textsf{Channel::getHeadTransform}.

For head-mounted displays (HMD), the tracking information is used to
move the frusta with the observer, as shown in \fig{fHMD}. This results
in different projections compared to normal tracking with fixed
projection screens.

@@

{\footnotesize\lstinputlisting[firstline=59, lastline=64, xleftmargin=-7em]
  {../../../src/examples/eqPly/view.h}}
{\footnotesize\lstinputlisting[firstline=58, lastline=76]
  {../../../src/examples/eqPly/view.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=40, lastline=47]
  {../../../src/examples/eqPly/view.cpp}}
{\footnotesize\lstinputlisting[firstline=78, lastline=85]
  {../../../src/examples/eqPly/view.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=547, lastline=570]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=491, lastline=533]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=305, lastline=337]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=346, lastline=371]
  {../../../src/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=87, lastline=98]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=98, lastline=108]
  {../../../website/documents/design/fileFormat.shtml}}
{\footnotesize\lstinputlisting[firstline=172, lastline=172]
  {../../../website/documents/design/fileFormat.shtml}}
{\footnotesize\lstinputlisting[firstline=178, lastline=178]
  {../../../website/documents/design/fileFormat.shtml}}
{\footnotesize\lstinputlisting[firstline=205, lastline=205]
  {../../../website/documents/design/fileFormat.shtml}}
{\footnotesize\lstinputlisting[firstline=249, lastline=249]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=107, lastline=125]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=125, lastline=136]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=136, lastline=158]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=158, lastline=167]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=172, lastline=176]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=178, lastline=181]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=181, lastline=202]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=205, lastline=225]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=225, lastline=246]
  {../../../website/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=249, lastline=333]
  {../../../website/documents/design/fileFormat.shtml}}

>>>>

\date{
  \vspace{14cm}
\begin{textblock}{}(0,4.5)
  \hspace{-1cm}\includegraphics[width=22cm]{images/teaser.pdf}
\end{textblock} {\Large The official reference for developing and
  deploying parallel, scalable OpenGL\texttrademark\ applications using
  the Equalizer parallel rendering framework}\\\vspace{1cm}
Version 1.8.2 for Equalizer 1.0+git \\[\medskipamount]
%  \today
}

@@

\clearpage
\pagenumbering{roman}
\pagestyle{headings}
\tableofcontents
%\clearpage
\listoffigures \vfill
\section*{Revision History}
{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Revision for Equalizer 0.5\\
    1.4     & Nov 25, 2008 & Revision for Equalizer 0.6\\
    1.6     & Aug 07, 2009 & Revision for Equalizer 0.9\\
    1.8     & Mar 21, 2011 & Revision for Equalizer 1.0\\
    1.8.1   & Mar 25, 2011 & Minor Revision\\
    1.8.2   & Apr 04, 2011 & Added \sref{sFocusDistance}\\
    \hline
  \end{tabularx}}
\clearpage

@@

{\footnotesize\lstinputlisting[firstline=107, lastline=123]
  {../../../Equalizer/examples/eqHello/hello.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=57, lastline=66]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=40, lastline=55]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=67, lastline=70]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=71, lastline=78]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=80, lastline=81]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=83]
  {../../../Equalizer/examples/eqPly/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=80, lastline=88]
  {../../../Equalizer/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=90, lastline=100]
  {../../../Equalizer/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=101, lastline=118]
  {../../../Equalizer/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=120, lastline=147]
  {../../../Equalizer/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=148, lastline=152]
  {../../../Equalizer/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=154, lastline=165]
  {../../../Equalizer/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=167, lastline=175]
  {../../../Equalizer/examples/eqPly/eqPly.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=61, lastline=74]
  {../../../Equalizer/examples/eqPly/initData.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=122, lastline=129]
  {../../../Equalizer/examples/eqPly/frameData.h}}
{\footnotesize\lstinputlisting[firstline=54, lastline=80]
  {../../../Equalizer/examples/eqPly/frameData.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=143, lastline=174]
  {../../../Equalizer/examples/eqPly/vertexBufferDist.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=74, lastline=95]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=97, lastline=99]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=100, lastline=118]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=130, lastline=139]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=310, lastline=317]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=319, lastline=327]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=329, lastline=359]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=387, lastline=390]
  {../../../Equalizer/examples/eqPly/config.cpp}}
{\footnotesize\lstinputlisting[firstline=450, lastline=481]
  {../../../Equalizer/examples/eqPly/config.cpp}}
{\footnotesize\lstinputlisting[firstline=528, lastline=534]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=284, lastline=308]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=790, lastline=859]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=35, lastline=48]
  {../../../Equalizer/examples/eqPly/node.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=257, lastline=268]
  {../../../Equalizer/examples/eqPly/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=57, lastline=70]
  {../../../Equalizer/examples/eqPly/pipe.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=72, lastline=78]
  {../../../Equalizer/examples/eqPly/pipe.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=39, lastline=55]
  {../../../Equalizer/examples/eqPly/pipe.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=80, lastline=84]
  {../../../Equalizer/examples/eqPly/pipe.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=76, lastline=98]
  {../../../Equalizer/examples/eqPly/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=100, lastline=109]
  {../../../Equalizer/examples/eqPly/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=69, lastline=78]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=123, lastline=137]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=139, lastline=165]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=167, lastline=178]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=591, lastline=610]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=619, lastline=689]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=238, lastline=261]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=36, lastline=57]
  {../../../Equalizer/examples/eqPixelBench/configEvent.h}}

@@

{\footnotesize\lstinputlisting[firstline=58, lastline=71]
  {../../../Equalizer/examples/eqPixelBench/config.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=55, lastline=61]
  {../../../Equalizer/examples/eVolve/main.cpp}}
...
{\footnotesize\lstinputlisting[firstline=92, lastline=95]
  {../../../Equalizer/examples/eVolve/main.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=36, lastline=44]
  {../../../Equalizer/examples/eVolve/error.h}}

@@

{\footnotesize\lstinputlisting[firstline=45, lastline=57]
  {../../../Equalizer/examples/eVolve/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=37, lastline=73]
  {../../../Equalizer/examples/eVolve/error.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=37, lastline=43]
  {../../../Equalizer/examples/eVolve/window.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=124, lastline=140]
  {../../../Equalizer/libs/client/aglWindow.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=67, lastline=101]
  {../../../Equalizer/libs/client/glXWindow.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=178, lastline=224]
  {../../../Equalizer/libs/client/wglWindow.cpp}}

\subsection{\label{sTracking}Stereo and Immersive Rendering}

Equalizer fully supports immersive rendering for Virtual Reality, supporting
multiple tracked observers, head-mounted displays (HMD) and flexible focus
distance support. Equalizer does not support tracking devices directly. Other
open source and commercial implementations cover this functionality well
and can easily be integrated with Equalizer.

An Equalizer configuration contains a number of observers. Each observer
represents on tracked entity. Most immersive configurations only have one
observer. It is possible to support multiple tracked viewers in the same
configuration, e.g., to use two HMDs.

Each observer has its own head matrix, eye separation and focus information for
tracking. In \textsf{eqPly}, all observers are updated with the same data. A
typical application would assign different tracking data to different observers.

The \textsf{eqPly} example contains basic support for head tracking to
illustrate how applications use these features.

\subsubsection{Stereo Rendering}

@@

\begin{figure}[h!t]\center
  \subfigure[]{\label{fMono}
    \includegraphics[width=.382\textwidth]{images/mono}}\hfil
  \subfigure[]{\label{fStereo}
    \includegraphics[width=.382\textwidth]{images/stereo}}
  {\caption{Monoscopic (a) and Stereoscopic (b) Frusta}}
\end{figure}

In stereo rendering, the scene is rendered twice, with the two frusta
'moved' by the distance between the eyes, as shown in \fig{fStereo}. The stereo
frusta are asymmetric, offset by half of the eye base. The stereo convergence
plane is the same as the projection surface, unless specified otherwise using
the focus distance API (see \sref{sFocusDistance}).

Note that while stereo rendering is often implemented the simpler toe-in method,
Equalizer uses the correct approach using asymmetric frusta.

\subsubsection{Immersive Rendering}

In immersive visualization, the observer is tracked in and the view
frusta are adapted to the viewer's position and orientation, as shown
in \fig{fTracked}. The transformation $origin \rightarrow viewer$ is set by
the application using \textsf{Config::setHead\-Matrix}, which is used by
the server to compute the eye positions, and consequently the view frusta.

\begin{figure}[h!t]\center
  \subfigure[]{\label{fTracked}
    \includegraphics[width=.382\textwidth]{images/tracked}}\hfil
  \subfigure[]{\label{fHMD}
    \includegraphics[width=.382\textwidth]{images/hmd}}
  {\caption{\label{fImmersive}Tracked (a) and HMD (b) Immersive Frusta}}
\end{figure}

Projection surfaces which are not X/Y-planar create frusta which are not
oriented along the Z axis, as shown in \fig{fFixed}. These frusta are positioned
using the channel's head transformation, which can be retrieved using
\textsf{Channel::getHeadTransform}.

For head-mounted displays (HMD), the tracking information is used to
move the frusta with the observer, as shown in \fig{fHMD}. This results
in different projections compared to normal tracking with fixed
projection screens. These difference is transparent to Equalizer applications,
only the configuration file uses a different wall type for HMD's.

\subsubsection{\label{sFocusDistance}Focus Distance}

The focus distance, also called stereo convergence, is the Z distance in which
the left and right eye frusta converge. A plane parallel to the near plane
positioned in the focus distance creates the same 2D image for both eyes.

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/focusFixed}
  {\caption{\label{fFixed}Fixed focus distance}}
\end{wrapfigure}
The frustum calculation up to Equalizer version 1.0 places the stereo
convergence on the projection surface, as shown in \fig{fFixed}. The focus
distance is therefore the distance between the origin and the middle of the
projection surface.

In Equalizer 1.1 and later the focus distance and mode can be configured and
changed at runtime for each \textsf{observer} and on the \textsf{config} for
untracked views.

The default focus mode \textsf{fixed} implements the algorithm used by Equalizer
1.0. This mode ignores the focus distance setting. The focus modes
\textsf{relative\_to\_origin} and \textsf{relative\_to\_observer} use the focus
distance to dynamically change the stereo convergence.

\begin{figure}[h!t]\center
  \subfigure[]{\label{fOrigin}
    \includegraphics[width=.49\textwidth]{images/focusOrigin}}\hfil
  \subfigure[]{\label{fObserver}
    \includegraphics[width=.49\textwidth]{images/focusObserver}}
  {\caption{\label{fImmersive}Dynamic focus distance relative to origin
      (a) and observer (b)}}
\end{figure}

The focus distance calculation relative to the origin allows to change the
focus independent of the observer position by separating the projection surface
from the stereo convergence plane upon which the observer focuses itself. The
convergence plane of the first wall in the negative Z direction is moved to be
at the given focus distance, as shown in \fig{fOrigin}. All other walls are
moved by the same relative amount. The movement is made from the view of the
central eye, thus leaving the mono frustum unchanged. \fig{fOrigin} shows the
new logical 'walls' used for frustum calculations in white, while the physical
projection from \fig{fFixed} is still visible in the background.

The focus distance calculation relative to the observer is similar to the origin
algorithm, but it keeps the closest wall in the observer's view direction at the
given focus distance, as shown in \fig{fObserver}. When the observer moves
forward, the focal plane moves forward as well. Consequently, when the observer
looks in a different direction, a different object in the scene is focused, as
indictated by the dotted circle in \fig{fObserver}.

\begin{figure}[h!t]\center
  \subfigure[]{\label{fFixedEx}
    \includegraphics[height=6.3cm]{images/focusFixedEx}}\hfil
  \subfigure[]{\label{fOriginEx}
    \includegraphics[height=6.3cm]{images/focusOriginEx}}\hfil
  \subfigure[]{\label{fObserverEx}
    \includegraphics[height=6.3cm]{images/focusObserverEx}}
            {\caption{\label{fImmersiveEx}Fixed(a), relative to origin(b) and
                observer(c) focus distance examples}}
\end{figure}

\fig{fImmersiveEx} shows an example for the three focus distance modes. The
configured wall is one meter behind the origin, the model two meters behind and
the observer is half a meter in front of the origin. The focus distance was set
to two meters.

@@

{\footnotesize\lstinputlisting[firstline=59, lastline=64, xleftmargin=-7em]
  {../../../Equalizer/examples/eqPly/view.h}}
{\footnotesize\lstinputlisting[firstline=58, lastline=76]
  {../../../Equalizer/examples/eqPly/view.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=40, lastline=47]
  {../../../Equalizer/examples/eqPly/view.cpp}}
{\footnotesize\lstinputlisting[firstline=78, lastline=85]
  {../../../Equalizer/examples/eqPly/view.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=547, lastline=570]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=491, lastline=533]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=305, lastline=337]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=346, lastline=371]
  {../../../Equalizer/examples/eqPly/channel.cpp}}

@@

{\footnotesize\lstinputlisting[firstline=87, lastline=98]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=98, lastline=108]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
{\footnotesize\lstinputlisting[firstline=172, lastline=172]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
{\footnotesize\lstinputlisting[firstline=178, lastline=178]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
{\footnotesize\lstinputlisting[firstline=205, lastline=205]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
{\footnotesize\lstinputlisting[firstline=249, lastline=249]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=107, lastline=125]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=125, lastline=136]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=136, lastline=158]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=158, lastline=167]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=172, lastline=176]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=178, lastline=181]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=181, lastline=202]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=205, lastline=225]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=225, lastline=246]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}

@@

{\footnotesize\lstinputlisting[firstline=249, lastline=333]
  {../../../equalizergraphics.com/documents/design/fileFormat.shtml}}
